---
format: gfm
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggtime <img src="man/figures/logo.svg" align="right" height="139" alt="ggtime website" />

```{=html}
<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/ggtime)](https://CRAN.R-project.org/package=ggtime)
<!-- badges: end -->
```

The ggtime package extends the capabilities of 'ggplot2' by providing 
grammatical elements and plot helpers designed for visualizing time series 
patterns. These functions use calendar structures implemented in the mixtime 
package to help explore common time series patterns including trend, 
seasonality, cycles, and holidays.

The plot helper functions make use of the tsibble data format in order to 
quickly and easily produce common time series plots. These plots can also be
constructed with the underlying grammar elements, which allows greater 
flexibility in producing custom time series visualisations.

## Installation

You can install the **stable** version from
[CRAN](https://cran.r-project.org/package=ggtime):

``` r
install.packages("ggtime")
```

You can install the development version of ggtime locally with:

```r
# install.packages("remotes")
# "ggtime" is should match the folder of the ggtime source code on your computer
remotes::install_local("ggtime")
```

## Examples

The simplest time series visualisation is the time plot, which shows time continuously on the x-axis with the measured variables on the y-axis. A time plot is useful for identifying patterns that persist over a long period of time, such as trends and seasonality. A time plot can be created from a tsibble with the `autoplot()` helper function.

```{r timeplot}
#| warning: false
#| message: false
library(ggtime)
library(ggplot2)
library(tsibble)

tsibbledata::aus_production |>
  autoplot(Beer)
```

A time plot can also be constructed with the underlying grammar elements, which allows greater flexibility in the style and appearance of the plot. The `geom_time_line()` element is a time-aware extension of `ggplot2::geom_line()` that ensures line slopes accurately represent rates of change even over timezone changes, gaps, and duplicated time points.

```{r}
tsibbledata::aus_production |>
  ggplot(aes(x = Quarter, y = Beer)) +
  geom_time_line(colour = "steelblue")
```

To view the shape of the annual seasonal pattern, it is useful to use a seasonal plot which shows time cyclically on the x-axis. This makes it easier to identify the peaks, troughs, and overall shape of the seasonality. A seasonal plot is created from a tsibble with the `gg_season()` plot helper function.

```{r seasonplot}
tsibbledata::aus_production |>
  gg_season(Beer)
```
