---
title: "How `coord_loop()` works"
subtitle: "Cut and fold, rather than draw and clip"
format:
  html:
    toc: true
    toc-depth: 3
    fig-width: 7
    fig-height: 4
    code-fold: false
---

```{r setup, include = FALSE}
# Rendered against the source tree rather than an installed build, so that the
# internal helpers shown below are the ones in `R/`.
knitr::opts_knit$set(root.dir = normalizePath(".."))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 150
)
```

```{r}
#| message: false
library(ggplot2)
pkgload::load_all(".", quiet = TRUE) # or library(ggtime)
```

## The idea

A looped time axis shows several stretches of time superimposed on the same
window: twelve months of every year drawn on top of each other, or seven days
of every week.

There are two ways to get there.

**Draw and clip** (the previous implementation). Draw the whole panel once per
loop, translating each copy so that the loop of interest lands in the window,
and attach a clipping path so only that loop is visible. Correct, but the cost
is `O(n_loops × panel content)`, and clipping paths need R >= 4.2 and a cairo,
pdf or quartz device. Annotations spanning the panel (`geom_vline()`,
`annotate()` with `Inf`) get clipped away with everything else, and curved
coords need alignment hacks to look right.

**Cut and fold** (the current implementation). Move the *data* instead of the
drawing. Fold every time value into the first loop's window, cut any geometry
that crosses a loop boundary into one piece per loop, and draw the panel once.
Cost is independent of the number of loops, no clipping is involved, and
annotations are simply never cut.

The whole feature is that second idea plus the bookkeeping needed to make ggplot2
agree with it.

## The pipeline

```
setup_panel_params()   work out the cut points; re-window on loop 1
        │
        ▼
distance()             called only for connected geoms → sets a flag
        │
        ▼
transform()            cut ── fold ── rekey ── parent transform ── arrange
```

### 1. Cut points, and the drawn window

`setup_panel_params()` runs the parent coord's method twice. The first pass has
nothing to do with looping: it exists so that scale limits, user `xlim`/`ylim`
and expansion are all resolved before we decide where the cuts go.

```{r}
#| eval: false
uncut_params <- ggproto_parent(coord, self)$setup_panel_params(
  scale_x,
  scale_y,
  params
)
time_cuts <- loop_cuts(
  uncut_params,
  self$time_scale,
  self$loops,
  self$time_loops
)
```

The second pass re-runs with limits narrowed to the window actually drawn —
from the first cut, spanning the *longest* loop:

```{r}
#| eval: false
self$limits[[self$time_scale]] <- c(
  time_cuts[1],
  time_cuts[1] + max(diff(time_cuts))
)
```

`max(diff(time_cuts))` matters because loops need not be equal: months are
28–31 days long, years 365 or 366. Sizing the window to the longest loop means
no loop overflows it. Breaks, expansion and axis labels are then computed for
that window, so the axis describes one loop rather than the whole series.

Cuts are stored twice — as time values (`time_cuts`, for anything that needs to
talk about calendars) and as numbers in transformed data space (`loop_cuts`,
which is the space `transform()` actually receives).

### 2. Telling connected data from pointwise data

This is the one genuinely awkward part. A point cloud must be folded row by row;
a line must be folded *and* cut, with new vertices interpolated at each
boundary. Nothing in the `Coord` API says which kind of data is arriving.

The signal comes from `ggplot2:::coord_munch()`, which calls `coord$distance()`
immediately before `coord$transform()` — and is only used by geoms whose rows are
vertices of a connected shape. Pointwise geoms reach `transform()` directly. So
`distance()` sets a flag that `transform()` reads and clears:

```{r}
#| eval: false
distance <- function(self, x, y, panel_params) {
  self$munch_connected <- TRUE
  ...
}

transform <- function(self, data, panel_params) {
  connected <- isTRUE(self$munch_connected)
  self$munch_connected <- FALSE
  cut <- if (connected) cut_connected else cut_pointwise
  ...
}
```

Munching only happens because `is_linear()` returns `FALSE`. That single return
value buys four things at once:

1. `coord_munch()` is used, giving us the connectivity signal above.
2. `Inf` and `-Inf` are resolved against `backtransform_range()`, so panel-spanning
   annotations arrive as real coordinates instead of infinities.
3. `GeomRect` becomes `GeomPolygon` and `GeomSegment` becomes `GeomPath`, so
   rects, tiles, bars and columns all arrive as rings and are cut by the same
   code path as lines.
4. `GeomRaster` falls back to `geom_rect()` with a message — the one geom that
   loses out.

When the wrapped coord is linear there is no real munching to do, so
`distance()` returns zeros; a distance of zero gives `extra = 1`, and
interpolating a segment into one piece returns it unchanged.

::: {.callout-important}
## Do not fold before measuring distance

When the wrapped coord *does* need to munch (`coord_radial()`), `distance()`
delegates to the parent using **unfolded** time values.

Folding first seems natural — it keeps distances small — but it reports a
segment spanning a whole loop as having travelled nowhere. `munch_data()` then
computes `extra = 1` and never interpolates, so every arc is drawn as a straight
chord. The symptom is a radial plot that renders as a near-empty sliver.

The unfolded distance is the right measure: a segment crossing N loops becomes N
pieces, and each piece needs its own vertices.
:::

### 3. Cutting

Everything in `R/loop-cut.R` is a plain function on a data frame, with no
ggproto involved. That keeps it unit testable and lets the same code serve
cartesian and radial coords, since both hand `transform()` the same
scale-transformed data space.

`loop_index()` assigns each time value to a loop, clamping so that out-of-range
values and infinities land in the first or last loop rather than off the end:

```{r}
cuts <- c(0, 10, 20, 30) # three loops of width 10
ggtime:::loop_index(c(0, 5, 15, 25, -Inf, Inf, NA), cuts)
```

`fold_time()` translates a value into the first window, `t - cuts[loop] + cuts[1]`:

```{r}
ggtime:::fold_time(c(5, 15, 25), c(1, 2, 3), cuts)
```

`cut_pointwise()` applies that to every positional aesthetic on the time axis
(`x`, `xmin`, `xmax`, `xend`, ...) independently, adding no rows:

```{r}
ggtime:::cut_pointwise(
  data.frame(x = 15, xmin = 12, xmax = 25, y = 1),
  "x",
  cuts
)
```

`cut_connected()` is the interesting one. Wherever consecutive vertices of the
same path fall in different loops, it interpolates a *pair* of vertices at each
boundary crossed — one closing the piece in the old loop at the end of the
window, one opening the next piece at the start:

```{r}
ggtime:::cut_connected(
  data.frame(x = c(5, 15), y = c(1, 3), group = 1L),
  "x",
  cuts
)
```

A single segment can cross several boundaries at once, and each is inserted:

```{r}
ggtime:::cut_connected(
  data.frame(x = c(2, 28), y = c(0, 26), group = 1L),
  "x",
  cuts
)
```

The insertion is vectorised — one pass computing how many boundaries each
segment crosses, then a single `vctrs::vec_slice()` to expand the frame — so
cost is linear in vertices regardless of loop count.

### 4. Rekeying, and why rings survive

Pieces are rekeyed by `(original key, loop)`. This is what makes closed shapes
work without any polygon clipping algorithm.

`GeomRibbon$draw_group()` munches its upper edge left-to-right and its lower
edge right-to-left as two *separate* `x`/`y`/`id` frames, then reassembles them
with `polygonGrob(id = )`. Because the key is computed arithmetically from the
original key and the loop index — never from order of appearance — upper piece
*k* and lower piece *k* get the same id, and each loop closes into exactly one
polygon:

```{r}
upper <- ggtime:::cut_connected(
  data.frame(x = c(2, 28), y = c(5, 6), id = 1),
  "x",
  cuts
)
lower <- ggtime:::cut_connected(
  data.frame(x = c(28, 2), y = c(1, 0), id = 1),
  "x",
  cuts
)
identical(sort(unique(upper$id)), sort(unique(lower$id)))
```

Ids are also spaced apart by a stride, because `GeomRibbon` offsets the lower
edge's ids by `max(ids)` before drawing the outline. With contiguous ids, lower
piece *k* would collide with upper piece *k + max(ids)* and draw a spurious
connector across the panel:

```{r}
any(upper$id %in% (lower$id + 1))
```

This holds for any ring monotone along the time axis, which covers ribbons,
areas, rects, tiles, bars and columns. Non-monotone concave rings are a known
limitation.

### 5. Panel decoration must bypass all of this

Gridlines and axis keys pass through `Coord$transform()` exactly like layer data
does, but they describe the loop *window* and are already expressed in its
coordinates. Cutting them is actively wrong: the drawn window spans the longest
loop, so a break sitting on the next loop's start gets folded back onto the
panel's left edge, landing on top of the first label.

Three methods are wrapped to suppress cutting and layout:

```{r}
#| eval: false
as_decoration <- function(coord, expr) {
  old <- coord$is_decoration
  coord$is_decoration <- TRUE
  on.exit(coord$is_decoration <- old, add = TRUE)
  expr
}
```

applied to `render_bg()`, `render_fg()` and `train_panel_guides()` — the only
paths ggplot2 routes decoration through.

### 6. `arrange_loops()`: the seam for `coord_calendar()`

After cutting and the parent's transform, each row carries a `.loop` index.
What happens next is the only difference between looping and a calendar:

```{r}
#| eval: false
# coord_loop(): superimpose, so just drop the index
arrange_loops <- function(self, data, panel_params) {
  data$.loop <- NULL
  data
}

# coord_calendar(): loop k becomes row k, an affine squeeze on the other axis
arrange_loops <- function(self, data, panel_params) {
  row <- data$.loop
  data$.loop <- NULL
  n_row <- self$.n_row
  offset <- (n_row - row) / n_row
  stacked <- if (self$is_flipped) ggplot_global$x_aes else ggplot_global$y_aes
  for (col in intersect(stacked, names(data))) {
    data[[col]] <- data[[col]] / n_row + offset
  }
  data
}
```

`coord_calendar()` inherits all of the cutting and folding, and adds only the
row layout plus replication of the decoration grob down the panel.

Support for a new base coord is likewise a declaration rather than an
implementation. `specialize_coord_loop()` methods only name which panel param
holds time:

```{r}
#| eval: false
specialize_coord_loop.CoordCartesian <- function(coord, ...) {
  ggproto("CoordLoopCartesian", coord, time_scale = coord$time)
}
```

## It working

### Seasonal patterns

```{r}
#| fig-cap: "Yearly loops of monthly US accidental deaths."
uad <- tsibble::as_tsibble(USAccDeaths)
uad$index <- as.Date(uad$index)

p <- ggplot(uad, aes(x = index, y = value)) +
  geom_line()

p + coord_loop(time_loops = "1 year")
```

Because time stays continuous rather than being discretised into a factor, the
line connecting December to January is preserved — the segment leaving December
is cut at the year boundary and its remainder reappears at the left edge.

Loop points can also be given explicitly rather than as a duration:

```{r}
#| fig-cap: "Explicit loop points, at three arbitrary dates."
p + coord_loop(loops = as.Date(c("1973-01-01", "1974-07-01", "1977-01-01")))
```

### Annotations spanning the panel

Previously these were clipped away with the rest of the panel copy
([#11](https://github.com/mitchelloharawild/ggtime/issues/11)). They are
pointwise data, resolved against the backtransformed range and folded, never
cut:

```{r}
#| fig-cap: "hline, vline and an Inf-spanning segment, all drawn in full."
p +
  geom_hline(yintercept = 9000, colour = "firebrick") +
  geom_vline(xintercept = as.Date("1973-07-01"), colour = "steelblue") +
  annotate(
    "segment",
    x = -Inf,
    xend = Inf,
    y = 7500,
    yend = 10500,
    colour = "grey40",
    linetype = "dashed"
  ) +
  coord_loop(time_loops = "1 year")
```

### Rings crossing boundaries

Ribbons, areas and rects are all rings, and all cut by the same code:

```{r}
#| fig-cap: "A ribbon cut into one closed polygon per loop."
ggplot(uad, aes(x = index)) +
  geom_ribbon(
    aes(ymin = value - 800, ymax = value + 800),
    fill = "steelblue",
    alpha = 0.4,
    colour = "steelblue4"
  ) +
  geom_line(aes(y = value)) +
  coord_loop(time_loops = "1 year")
```

```{r}
#| fig-cap: "Rects spanning a loop boundary are split, not dropped."
quarters <- data.frame(
  xmin = as.Date(c("1973-11-15", "1974-11-15", "1975-11-15")),
  xmax = as.Date(c("1974-02-15", "1975-02-15", "1976-02-15"))
)

p +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    data = quarters,
    inherit.aes = FALSE,
    fill = "orange",
    alpha = 0.3
  ) +
  coord_loop(time_loops = "1 year")
ggplot(uad, aes(x = index, y = value)) +
  geom_line() +
  geom_rect(
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    data = quarters,
    inherit.aes = FALSE,
    fill = "orange",
    alpha = 0.3
  ) +
  coord_calendar(time_rows = "1 year")
```

### Radial

`coord_radial()` now draws genuine arcs across loop boundaries, with no
transparency or alignment workarounds:

```{r}
#| fig-width: 5
#| fig-height: 5
#| fig-cap: "A polar seasonal plot."
p + coord_loop(time_loops = "1 year", coord = coord_radial())
```

(The label at the top of the circle is doubled: the first and last break of the
window sit at the same angle once the axis wraps, so both are drawn. Cosmetic,
and specific to radial.)

### Calendars

The same machinery, with the pieces stacked into rows instead of superimposed:

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "Weekly rows of half-hourly pedestrian counts."
tsibble::pedestrian |>
  dplyr::filter(Date < "2015-02-01") |>
  ggplot(aes(x = Date_Time, y = Count, colour = Sensor)) +
  geom_line() +
  coord_calendar(time_rows = "1 week") +
  scale_x_datetime(date_breaks = "1 day", date_labels = "%a") +
  theme(legend.position = "bottom")
```

### Cost is flat in the number of loops

Drawing once means the number of loops no longer drives the cost. Twenty years
of daily data, rendered to a PNG:

```{r}
#| cache: true
daily <- data.frame(
  date = seq(as.Date("2000-01-01"), as.Date("2019-12-31"), by = "1 day")
)
daily$value <- sin(seq_len(nrow(daily)) / 58) + rnorm(nrow(daily), sd = 0.2)

render <- function(plot) {
  f <- tempfile(fileext = ".png")
  png(f, width = 800, height = 500)
  on.exit(
    {
      dev.off()
      unlink(f)
    },
    add = TRUE
  )
  print(plot)
}

timings <- vapply(
  c("1 year", "1 month", "1 week"),
  function(period) {
    gg <- ggplot(daily, aes(date, value)) +
      geom_line() +
      coord_loop(time_loops = period)
    system.time(render(gg))[["elapsed"]]
  },
  numeric(1)
)

round(timings, 2)
```

Roughly 20, 240 and 1040 loops respectively, in about the same time each. The
same three plots measured against the previous implementation (separately, since
it is no longer in the tree) took 1.37s, 0.85s and 1.87s — cost tracked loop
count, and the weekly case was the slowest by a wide margin.

### No device requirement

Clipping paths needed R >= 4.2 and a cairo, pdf or quartz device. Nothing here
does, which is easiest to show on a device that has never supported them:

```{r}
f <- tempfile(fileext = ".png")
png(f, type = "Xlib", width = 400, height = 300)
print(p + coord_loop(time_loops = "1 year"))
dev.off()
file.size(f) > 0
```

## Testing

Because the engine is pure functions, most of the behaviour is covered by fast
unit tests in `tests/testthat/test-loop-cut.R` rather than SVG snapshots —
boundary interpolation, multi-loop segments, ribbon key agreement, the id
stride, irregular (monthly) cut spacing, `NA`/`Inf` handling, and empty input.
Snapshots in `test-coord-loop.R` are left to cover what only shows up in a
rendered panel.

## Known limitations

All four are reachable from ordinary plots. The examples below are the shapes
they take in practice.

### Rings that re-enter the same loop

Rekeying by `(key, loop)` gives each loop exactly one piece of a ring — which is
right as long as the ring crosses each loop window once. A ring that doubles back
in time hands the same loop two disconnected fragments, and they are closed into
a single polygon that bridges the gap.

Monotone rings are safe, and that covers every ribbon, area, rect, tile, bar and
column. What is not safe is a hand-traced `geom_polygon()` whose outline reverses
direction — an arrow or chevron marking a period, say:

```{r}
#| fig-cap: "Unlooped, the chevron's notch is drawn as intended."
chevron <- data.frame(
  x = as.Date(c(
    "1973-08-01",
    "1973-11-01",
    "1973-08-01",
    "1974-04-01",
    "1974-07-01",
    "1974-04-01"
  )),
  y = c(7200, 8200, 9200, 9200, 8200, 7200)
)

marked <- p +
  geom_polygon(
    aes(x, y),
    data = chevron,
    inherit.aes = FALSE,
    fill = "orange",
    alpha = 0.5,
    colour = "darkorange3"
  )

marked
```

```{r}
#| fig-cap: "Looped, the notch's fragments are joined into stray wedges."
marked + coord_loop(time_loops = "1 year")
```

The two edges of the notch fall in the same loop but are not contiguous, so the
polygon closes across the panel. Drawing the shape as two monotone polygons
instead of one avoids it.

### Calendar rows are shared across panels

The row count is a maximum across panels, so every facet gets the same number of
rows. That is deliberate — a calendar month should be the same height in every
facet — but a panel covering less time than its neighbours is padded with empty
rows. A partial final year is the usual way to meet this:

```{r}
#| fig-width: 7
#| fig-height: 5
#| fig-cap: "2023 is only four months long, but is laid out with twelve rows."
daily_years <- data.frame(
  date = seq(as.Date("2021-01-01"), as.Date("2023-04-30"), by = "1 day")
)
daily_years$value <- sin(seq_len(nrow(daily_years)) / 30)
daily_years$year <- format(daily_years$date, "%Y")

by_year <- ggplot(daily_years, aes(date, value)) +
  geom_line() +
  facet_wrap(~year, scales = "free_x")

by_year + coord_calendar(time_rows = "1 month")
```

Eight empty rows under 2023. Because the count is derived by accumulating a
maximum, it is state on the coord rather than a property of a panel; it is reset
in `setup_params()` at the start of each build. The sharing itself is inherent
to a calendar layout: a common row height across facets is what makes the
calendar readable and comparable between panels.

### The piece ceiling

Rekeyed `id` values are spaced by a fixed `loop_id_stride` (`1e5`), because the
upper and lower edges of a ribbon are cut in separate calls and must agree on
keys without seeing each other's data. `polygonGrob()` then coerces `id` to
integer, which caps a single ribbon-like layer at `.Machine$integer.max /
loop_id_stride` — about 21,500 — pieces.

That is a lot of pieces, but loop counts multiply: 30 years of six-hourly data
folded into a daily profile is nearly eleven thousand loops, and a ribbon is cut
twice per loop.

```{r}
#| error: true
six_hourly <- data.frame(
  time = seq(
    as.POSIXct("1990-01-01", tz = "UTC"),
    by = "6 hours",
    length.out = 30 * 365 * 4
  )
)
six_hourly$value <- sin(seq_len(nrow(six_hourly)) / 13)

ggplot(six_hourly, aes(time, ymin = value - 0.2, ymax = value + 0.2)) +
  geom_ribbon(fill = "steelblue") +
  coord_loop(time_loops = "1 day")
```

An error rather than a silent mis-draw: past the ceiling `grid` would coerce the
ids to `NA` and quietly drop the geometry. The same plot over twenty years, or as
a `geom_line()` rather than a ribbon, is well inside the limit.
